Fast Bayesian Inference for Exoplanet Discovery in Radial Velocity Data

### 2014-12-13 Paper/ms.tex
More work in phase transition section

Even if the slab contains a very small amount of posterior mass, its existence
can still cause problems with the calculation of the marginal likelihood
if annealing methods are used. The thermodynamic integral formula gives
the log of the marginal likelihood as an average of log likelihoods:
\begin{eqnarray}
\log(\mathcal{Z}) &=& \int_0^1 \left<\log\left[L(\theta)\right]\right>_\beta \, d\beta
\end{eqnarray}
where the expectation is taken with respect to the
distribution with ``inverse temperature'' $\beta$, proportional to
$\pi(\theta)L(\theta)^\beta$. Even if the slab contains virtually zero mass
at $\beta=1$ (i.e. the posterior), at some inverse temperature between 0 and 1
the slab and the spike ought to contribute equally to the distribution. At
this critical temperature the MCMC will fail to mix and will give a misleadingly
low estimate of the average log likelihood at that temperature.

In data analysis problems, phase transitions can exist when the data contains
a ``big'' effect which provides a lot of information about some parameters, and
a ``small'' subtle effect.
Nested Sampling, and variants such as DNS, are not affected
by phase transitions. Part of their output, the relationship between the
likelihood $L$ and the enclosed prior mass $X$, can be used to diagnose whether the
problem contains a phase transition. In particular, if the graph of $\log(L)$
vs. $\log(X)$ is convex at some point, then a phase transition exists.

>>>>

If the slab contains a very small amount of posterior probability, it is not
a problem if an MCMC algorithm spends all its time in the spike. However, this
situation could still cause problems with the calculation of the marginal likelihood
if annealing methods are used. The thermodynamic integral formula gives
the log of the marginal likelihood $\mathcal{Z}$ as an average of log likelihoods:
\begin{eqnarray}
\log(\mathcal{Z}) &=& \int_0^1 \left<\log\left[L(\theta)\right]\right>_\beta \, d\beta
\end{eqnarray}
where the expectation is taken with respect to the
distribution with ``inverse temperature'' $\beta$, proportional to
$\pi(\theta)L(\theta)^\beta$. Even if the slab contains virtually zero probability
when $\beta=1$ (i.e. the posterior), for some values of the inverse temperature
$\beta$ the slab and the spike will both be important. At
these temperatures the MCMC will fail to mix (it will incorrectly spend all its
time in either the slab or the spike, rather than mixing between the two)
and will give a misleading estimate of the average log likelihood at that temperature.

In data analysis problems, phase transitions can exist when the data contains
a ``big'' effect which provides a lot of information about some parameters, and
a ``small'' subtle effect.
Nested Sampling, and variants such as DNS, are not affected
by phase transitions because the exploration only makes use of likelihood
{\it rankings}, rather than likelihood values themselves.
Part of their output, the relationship between the
likelihood $L$ and the enclosed prior mass
$X(L) = \int \pi(\theta) \mathds{1}\left[L(\theta) > L\right]\, d^n\theta$
can be used to diagnose whether the
problem contains a phase transition. In particular, if the graph of $\log(L)$
vs. $\log(X)$ is convex at some point, then a phase transition exists
\citep{skilling}.
